\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-05-09 doc/incertae_sedis.tex
shorter...

However, this assumes that \emph{incertae sedis} taxa are always
placed within their siblings, or descendants of their siblings.
As
such, it does not handle cases such as Figure \ref{fig:Placing-one-incertae}, 
in which $I$ is first placed as sibling to $AI$, and then
$AI$ is placed within $I$.
Note that $I$ is not a sibling of $AI$ or a
descendant of a sibling.
Furthermore, the MRCA of $I$ and $AI$ is the
root, and the path from $AI$ to the root includes $A$, but $A$ is not
broken.
We suggest that this conundrum may be resolved by first
processing the placement of $I$ within $A$ by modifying the taxonomy
to reflect this placement\footnote{Does this actually work? Some
placements break taxa.}.
When we consider the placement of $AI$ within
$I$ on this modified taxonomy, the MRCA of $I$ and $AI$ is now $A$
instead of the root, and the correct interpretation results.
We
suggest that by walking the tree in a pre-order fashion and processing
shallowing placements first, the previous approach of finding the MRCA
and labelling intermediate nodes as either additional placements or
broken incertae sedis (or incertae\_sedis\_inherited) taxa holds.
Such
an approach should also handle cases containing nested incertae sedis
taxa, as in Figure \ref{fig:Placing-one-incertae}.

\begin{comment}
One question is whether this approach is able to handle nested
incertae sedis taxa.

Placement of incertae sedis taxa by input trees is unfortunately not
quite as simple as finding a single location where an I.S.
taxon
should attach.
For example, when an incertae sedis taxon is broken,
its children need to be ``placed'' separately.

Each input tree can relate to an \emph{incertae sedis} taxon
$(A,B,C)D$ in a number of ways
\begin{itemize}
    \item it could resolve
$A$, $B$, $C$, or $D$.
    \item it could place $D$ on a degree-2 (=out-
degree-1) node that bisects a branch 
    \item it could place a descendant
taxon of $D$ 
    \item it could place a descendant taxon of $D$ in a
\emph{different place} than another input tree.
    \item it could place
children of $D$ in multiple places, thus conflicting with the branch.
If the \emph{incertae sedis} taxon $(A,B,C)$ is broken, then $A$, $B$
and $C$ become \emph{incertae sedis} clades in their own right, that
may attach separately, except that they .
This is because none of $A$,
$B$, or $C$ is in the exclude set of the siblings of $D$.
\end{itemize}
\end{comment}

\section{The rooted-split semantics in practice}

In order to examine the results of the rooted splits-semantics above, we explore
the outcome on a series of complex cases.  These cases are intended to
illustrate some of the issues that arise when developing an operational semantics of
\emph{incertae sedis}, as well as the outcomes for the rooted-split semantics.
%We think that they constitute one of the main contributions of the paper.
In each case, the synthesis trees in this section are constructed using the
BUILD-based algorithm described in section \ref{sec:background}, without
decomposing them into smaller subproblems or pruning and then unpruning
taxonomy-only taxa.  Synthesis trees that are described as \emph{incertae
sedis}-aware apply the BUILD algorithm to splits constructed using equation
\ref{eq:exsetformonea}, whereas the synthesis trees that are not \emph{incertae
sedis}-aware are constructed from splits using equation
\ref{eq:exsetformonea-traditional}.

Much of the complexity that we explore here comes from the fact that it is not
only tip taxa that can be \emph{incertae sedis}.  It is also possible to have
higher ranked  \emph{incertae sedis} taxa, such as genera that are marked
\emph{incertae familia}.

\subsection{The \emph{incertae sedis} semantics restores taxonomic groups and names}

With the naive semantics, intrusion of an \emph{incertae sedis} taxon into a sibling is
interpreted as conflict with that sibling.  This results in the loss of the edge
corresponding to the sibling taxon as well as the taxon name.

We see from Fig. \ref{fig:Handling-incertae-sedis}d that use of the unrestricted intrusion
semantics results the taxonomy edge $AB$ grouping $A$ and $B$.  It also restores the
names $AB$ and $A$.  We may thus determine that $C$ has been placed within $A$ (and
its ancestor $AB$).

\subsection{Conflict with \emph{incertae sedis} taxa}

When only tip nodes on the taxonomy are \emph{incertae sedis}, we have
a very simple form of the \emph{incertae sedis} supertree problem.
However, in practice entire clades may be \emph{incertae sedis}, and
so more complex issue arise.
In this section we consider a number of
cases that must be handled when attempting to place \emph{incertae
sedis} taxa.

\emph{Note that the synthesis of all input trees before the taxonomy
is unaffected by incertae sedis information.}

\subsection{Conflicting placement among input trees}

{[}\textbf{BDR:} \emph{maybe we should merge this with Figure.
\ref{fig:Broken-incertae-sedis}.
Although the point of figure
\ref{fig:placement-example1} is to talk about sub-problem
decomposition as well as what the result looks like.}{]}

The addition of \emph{incertae sedis} taxa allows new types of
conflict between input trees.
For example, different input trees might
place an incertae sedis taxon in conflicting locations.
This is
illustrated in Figure \ref{fig:placement-example1}, where the IS
taxon (ott7,ott6)ott10 is placed as sister to ott1 by phylogeny
$\tau_{1}$ and as sister ott3 by phylogeny $\tau_{2}$.

When this happens, the placement of the IS taxon is not influenced by
its being marked IS on the taxonomy.
Thus, in Example 1, the higher
ranked tree $\tau_{1}$ will be reflected in the synthesis tree, ott10
will be placed as sister to ott1.
In contrast, the conflicting
placement in $\tau_{2}$ will not be reflected in the synth tree.

All this would occur in the previous version of propinquity.
Where the
updated version differs that (a) the names ott9 and ott8 are retained
instead of being dropped.
(b) as a result of not breaking ott8 and
ott9, we do not move ott3 and ott2 up to ott11.
{[}BDR: Extend more figures!{]}

\subsection{Case 1: An \emph{incertae sedis} clade}

When a clade is marked as \emph{incertae sedis}, we need to consider
two cases.
In the first case, the clade may be placed within a sister
clade intact (Figure \ref{fig:Handling-incertae-sedis}).
However, if
the clade is not monophyletic in the synthesis tree, then we allow the
members of the clade to be placed separately (Figure \ref{fig:Broken-incertae-sedis}).

When the input phylogenies conflict on where members of an incertae
sedis clade should be placed, neither placement is rejected.
Instead,
the taxon $C$ is broken, the name $C$ disappears, and $C3$ floats to
the top level.
Algorithmic, alternatively placements of children of an
incertae sedis clade lead to contesting of all edges along with taxa
might be placed, creating a large, merged sub-problem, as in Figure
\ref{fig:placement-example1}.

In Figure \ref{fig:Broken-incertae-sedis}, tree $T_{1}$ places $T1$
next to $R1$, while $T_{2}$ places $T2$ next to $R2$.
Since different
members of $C$ are placed, neither placement for $T$ is rejected.
Instead the taxon $T$ is broken, the name $T$ disappears, and $T3$
floats to the MRCA of the placed members of $T$.
Furthermore, since taxon $T$ is a broken
\emph{incertae sedis} taxon, all of its children are effectively
incertae sedis independently, with the exception that they cannot be
placed within each other.
Therefore, the names $R$ and $RS$ are not
lost, since the taxa placed within them are \emph{incertae sedis}
names.

The situation here would be different if the monophyly of $T$ was
supported by an high-ranked input phylogeny.
In that case, instead of
breaking $T$, we would choose the placement in the highest-ranked
tree.

\subsection{Case 2: A nested \emph{incertae sedis} clade}

% \begin{figure}
% \subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-nested/tree3-shaded}

% }\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/is-nested/tax-shaded}

% }

% \subfloat[Synthesis tree (\emph{incertae sedis} ignored)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-no-is-shaded}

% }\hfill{}\subfloat[Synthesis tree (\emph{incertae sedis} considered)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-is-shaded}

% }

% \caption{\label{fig:Nested-incertae-sedis}Nested incertae sedis taxa.
% (a)
% Input tree \#1 places CI1 within CG, and C within $A$.
% (b) In the
% taxonomy, CI is \emph{incertae sedis }within C, while C is \emph{incertae
% sedis} under the root.
% (c) The synthesis tree when \emph{incertae
% sedis} taxa are not considered.
% The names AB, A, and CG are lost.
% (d) The synthesis tree when \emph{incertae sedis} taxa are considered.
% The names AB, A, and CG are regained, as well as the edges leading
% to AB and CG.}
% \end{figure}

When \emph{incertae sedis} clades are allowed, it is possible for an
\emph{incertae sedis} taxon to be nested within another \emph{incertae
sedis} taxon.
In such a case we must allow the higher-ranked taxon to
be placed within a sibling, while preserving the right of the lower
ranked taxon to be placed within a lower-level sibling (Figure
\ref{fig:Nested-incertae-sedis}).

\subsection{Case 3: Placement of one \emph{incertae sedis} taxon within
another.}
\subsection{Surprising results of the split-based semantics}

\subsubsection{Inter-digitation}
1. Inter-digitation

\bdrcomment{Ben: discuss figure \ref{fig:interdigitation}.}

\subsubsection{Placing one \emph{incertae sedis} group within another}
It would be nice if our semantics of incertae sedis taxa would allow us to place
one \emph{incertae sedis} sibling inside another.
A special case of
this is when a higher-rank incertae sedis taxon $A$ is placed as
sister to an incertae sedis taxon $B$, and then $B$ is placed within
$A$ (Figure \#).

2. Placing A next to B, then B within A.

\bdrcomment{Ben: Write discussion of Figure \ref{fig:Placing-one-incertae}}

\subsection{Splits-based semantics versus edit-based semantics}

Here we focus on a split-based semantics instead of a semantics based on applying
of a series of edit operations to the tree. The split-based semantics allows us
to seek trees that are simultaneously consistent with a set of split-constraints.
The use of a split-based semantics does impose some limitations. These come
primarily because in the edit-based semantics, after we place an
\emph{incertae sedis} taxon, it is no loner \emph{incertae sedis}, whereas with
the split-based semantics taxa are either always \emph{incertae sedis} or never
\emph{incertae sedis}.

This has an effect, for example, when we consider two incertae sedis siblings $A$
and $B$. If we can place $A$ within $B$, and $B$ within $A$, then a split- based
semantics allows us to freely inter-digitate $A$ and $B$. If we do not want to
allow $A$ and $B$ to be inter-digitated, then we cannot both $A$ within $B$ and $B$
within $A$. In addition, we can imagine that a family $A$ is incertae sedis within a
kingdom. In an edit-based semantics, we could place $A$ as a sibling to a genus $B$,
mark $A$ as non-\emph{incertae-sedis}, and then place $B$ within $A$. In a split-based
semantics, we must disallow either $B$ within $A$ or $A$ within $B$ if we want to disallow
interdigitation.

As a result of this, we cannot satisfy all five of these criteria. Thus, we
describe two alternative split-based semantics. The non-interdigitation
semantics gets \# (1,2,4,5), and the interdigitation semantics gets (1,3,4,5).
This makes the point that each semantics results from different choices about
which criteria to satisfy.

@@

In order to handle \emph{incertae sedis} taxa within propinquity, we
must modify some of the stages of the propinquity pipeline.
Subproblem
decomposition must place \emph{incertae sedis} taxa in the correct
subproblem.
Subproblem files must indicate which taxa are incertae
sedis.
The subproblem solver must read this information, account for
\emph{incertae sedis} taxa when solving subproblems, and correctly
name taxa that have been modified by having \emph{incertae sedis} taxa
place inside them.
The unpruner must be aware of \emph{incertae sedis}
taxa.
Annotations of the tree must be aware of \emph{incertae sedis}
taxa so that it does not consider taxa broken when they have an
incertae sedis taxon placed inside them.

\subsection{Exemplifying taxa}

One current problem is that well-known taxa like Fungi or Mammalia
tend to have a very large number of incertae sedis children, making
browsing in the tree viewer difficult.
This can happen when, for
example, fossils or other hard-to- place taxa get classified only to
the level of these well-known nodes and no further.
This leads to a
situation where well-known taxa serve as a dumping ground for unplaced
taxa.

Our current approach to this problem is to perform a second round of
pruning, or ``cleaning'', during the exemplification step.
Incertae
sedis taxa are pruned at this stage if they do not occur in any input
trees.
We thus generate a second ``cleaned taxonomy'' that has
undergone this further round of cleaning.
This approach improves on
the previous approach in that \emph{incertae sedis} taxa in input
trees are no longer pruned.
This approach also removes tons of
\emph{incertae sedis} children from nodes like ``Fungi'', where a lot
of unplaced fossils with few observable characters have been dumped.

However, this approach has the negative effect of pruning some
incertae sedis taxa that need not be pruned.
For example, suppose
\emph{incertae sedis} taxon $A$ contains 5 children, of which only 1
child $A_{1}$ occurs in an input tree.
If the taxon $A$ is not broken,
then it should be possible to attach the other 4 members of $A$ next
to $A_{1}$, without cluttering up the synthesis tree.
Such taxa have
been successfully placed even though they are not in any input tree.
This can only be discovered after synthesis is complete, though.

We should be able to to figure out which incertae sedis taxa are
placed (see section \ref{subsec:Placement}), and then mark the other
ones incertae sedis in the \emph{synthesis tree}.
We could then filter
out the taxa that \emph{remain }incertae sedis (or incertae\_sedis
inherited) in the tree viewer instead of the synthesis pipeline.

@@

\caption{\label{fig:placement-example1}Example.
An incertae sedis clade (ott6,ott7)
is placed in different subtrees by input trees $\tau_{1}$ and $\tau_{2}$.
In $\tau_{1}$, two nodes that correspond to the taxonomy their ingroup
extended to include (ott6,ott7), and the branches leading to these
nodes have been colored blue.
The dashed blue edge leads to a node
that is a newly-introduced degree-2 node which does not correspond
to any taxonomy node.
In $\tau_{2}$, only one node that corresponds
to a taxonomy node needs to have its ingroup extended.
The placement
of (ott6,ott7) into ott8 toward ott1 by $\tau_{1}$ conflicts with
the placement of (ott6,ott7) into ott9 toward ott3 by $\tau_{2},$}
\end{figure}
 For example, in Figure
\ref{fig:placement-example1}, input tree $\tau_{1}$ contests ott9 and
input tree $\tau_{2}$ contests ott9 and ott8.
Thus ott1, ott2, ott3,
ott6, and ott7 end up in the same sub-problem.

>>>>

\caption{\label{fig:placement-example1}Example.
    An incertae sedis clade (ott6,ott7)
    is placed in different subtrees by input trees $\tau_{1}$ and $\tau_{2}$.
    In $\tau_{1}$, two nodes that correspond to the taxonomy their ingroup
    extended to include (ott6,ott7), and the branches leading to these
    nodes have been colored blue.
    The dashed blue edge leads to a node
    that is a newly-introduced degree-2 node which does not correspond
    to any taxonomy node.
    In $\tau_{2}$, only one node that corresponds
    to a taxonomy node needs to have its ingroup extended.
    The placement
    of (ott6,ott7) into ott8 toward ott1 by $\tau_{1}$ conflicts with
    the placement of (ott6,ott7) into ott9 toward ott3 by $\tau_{2},$}
\end{figure}

@@

In order to handle \emph{incertae sedis} taxa within propinquity \citep{redelings2017supertree}, we
must modify some of the stages of the propinquity pipeline.
Subproblem
decomposition must place \emph{incertae sedis} taxa in the correct
subproblem.
Subproblem files must indicate which taxa are \emph{incertae sedis}.
The subproblem solver must read this information, account for
\emph{incertae sedis} taxa when solving subproblems, and correctly
name taxa that have been modified by having \emph{incertae sedis} taxa
place inside them.
The unpruner must be aware of \emph{incertae sedis}
taxa.
Annotations of the tree must be aware of \emph{incertae sedis}
taxa so that it does not consider taxa broken when they have an
\emph{incertae sedis} taxon placed inside them.

@@

 For example, in Figure
\ref{fig:placement-example1}, input tree $\tau_{1}$ contests ott9 and
input tree $\tau_{2}$ contests ott9 and ott8.
Thus ott1, ott2, ott3,
ott6, and ott7 end up in the same sub-problem.
